Verifying a Postfix Language \\ \large CS595 Final Project Proposal

### 2021-11-26 project/paper.tex
organize, start on typing section

\section{Preliminary Work - The Language}
This section describes some of the key features of the original language. It is included only to show that the subset is a good representation of the original language and thus can be skipped.

@@

\section{Prelimnary work - The Subset}
Here we describe at a high level the subset of the language to formalize.

@@

    e \bnfdef & e~e~+ \bnfalt ~e~e~\times \bnfalt ~ \kwtern{e}{e}{e} 
        \bnfalt e~e~== \bnfalt ~ \tau~ \tau~ ==
        \bnfalt ~ ( e^* )
        \bnfalt \kwmacro{\tau^*}{S} \bnfalt \kwfix{e}{\tau}
        \bnfalt \kwid{x} \bnfalt \kwunesc{x}\\

@@

    S \bnfdef & \varepsilon \bnfalt S~ S \bnfalt e~ \kweid{x} = \bnfalt ~e

@@

Notes
\begin{itemize}
    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$
    \item notice that unpack can have any non-negative number of output values depending on the given input
    \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
\end{itemize}

\section{Semantics}
We've expanded the macro value syntax to also capture lexically scoped variables. These are represented as with a brackets syntax as shown: $(( \tau^* )[ x \step (), y \step 123 ]: ... )$.

{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{Num Lits}
             {\forall ~ T \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
             {\kwnumlit{n}{T} \val}
        \and

        % Derivation of e* -> e*'
        \Rule{$e^* \step~ e^{*\prime}$ - Part 1}
            {e_1 \step e_1'}
            {e_1 ~ ... ~ e_n \step e_1' ~ ... ~ e_n}
        \and
        \Rule{$e^* \step~ e^{*\prime}$ - Part 2}
            {e_1 \step e_1'
                \\v_i \val ~ \forall v_i \in v*}
            {v* ~ e_1 ~ ... ~ e_n \step v* ~ e_1' ~ ... ~ e_n}
        \and

        % Tuple literals evaluation
        \Rule{Tuple Lits - 1}
            {e^* \step e^{*\prime}}
            {( e^* ) \step ( e^{*\prime} )}
        \Rule{Tuple Lits - 2}
            {v_i \val ~ \forall v_i \in v*}
            {(v*) \val}
        \and

        % unpack operator
        \Rule{Tuple Value Unpack - 1}
            {v_i \val ~ \forall i \in [1, n]}
            {\kwunpack{(v_1 ... v_n)} \step v_1 ... v_n}
        \Rule{Tuple Value Unpack - 2}
            {\strut}
            {\kwunpack{()} \step \varepsilon}
        \and
        \Rule{Tuple Type Unpack - 1}
            {\strut}
            {\kwunpack{(\tau_1 ... \tau_n)} \step \tau_1 ... \tau_n}
        \Rule{Tuple Type Unpack - 2}
            {\strut}
            {\kwunpack{\kwUnit} \step \varepsilon}

        % math
        \Rule{Numeric Sum}
            {v_1 \val \\ v_2 \val \\ v_3 = v_1 + v_2}
            {v_1~ v_2~ + \step v_3}
        \Rule{Numeric Product}
            {v_1 \val \\ v_2 \val \\ v_3 = v_1 * v_2}
            {v_1~ v_2~ \times \step v_3}
        \and

        % ternary
        \Rule{Ternary 1}
            {v_1 \val \\ v_2 \val}
            {\kwtern{\kwnumlit{0}{\kwInt}}{v_1}{v_2} \step v_2}
        \Rule{Ternary 2}
            {v_1 \val \\ v_2 \val \\ n \neq 0}
            {\kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2} \step v_1}

    \end{mathpar}
}

\end{document}

>>>>

\section{The Language}
This section describes some of the key features of the original language. It is included only to show that the subset is a good representation of the original language and thus can be skipped.

@@

\section{The Subset}
Here we describe at a high level the subset of the language to formalize.

@@

    e \bnfdef & e~e~+ \bnfalt ~e~e~\times \bnfalt ~ \kwtern{e}{e}{e} 
        \bnfalt e~e~== \bnfalt ~ \tau~ \tau~ ==
        \bnfalt ~ ( e^* )
        \bnfalt \kwmacro{\tau^*}{S} \bnfalt \kwfix{e}{\tau}
        \bnfalt \kwid{x} \bnfalt e ~ \textbf{@} 
        \bnfalt S~ S \bnfalt e~ \kweid{x} = \\

@@

@@

Notes
\begin{itemize}
    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$. It's possible that one could be implemented in the language itself.
    \item notice that unpack can have any non-negative number of output values depending on the given input
    \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
\end{itemize}

\section{Semantics}

% TODO we should split this into subsections
\subsection{Value Judgements}
Types come in the form of values and expressions just like other language features.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % number literals
        \Rule{V-1}
             {\forall ~ \tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
             {\kwnumlit{n}{\tau} \val}
        \and

        % type values
        \Rule{V-2}{\strut}{\kwInt \val} \and
        \Rule{V-3}{\strut}{\kwLong \val} \and
        \Rule{V-4}{\strut}{\kwFloat \val} \and
        \Rule{V-5}{\strut}{\kwDouble \val} \and
        \Rule{V-6}{\tau_1 \val \\ \tau_2 \val}{\tau_1 \tau_2 \texttt{|} \val} \and
        \Rule{V-7}{\tau_1 \val \\ \tau_2 \val}{\kwArrow{\tau_1}{\tau_2} \val} \and
        % fixed point combinator
        \Rule{Fixed Point}{e \val \\ \tau \val}{\kwfix{e}{\tau} \val}
    \end{mathpar}  
}

\subsection{Step Judgement}
Expressions are evaluated from left to right until everything is a value
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Derivation of e* -> e*'
        \Rule{STEP-1}
            {e_1 \step e_1'}
            {e_1 ~ ... ~ e_n \step e_1' ~ ... ~ e_n}
        \and
        \Rule{STEP-2}
            {e_1 \step e_1'
                \\v_i \val ~ \forall v_i \in v*}
            {v* ~ e_1 ~ ... ~ e_n \step v* ~ e_1' ~ ... ~ e_n}
        \and
    \end{mathpar}  
}

\subsection{Macros, Identifiers and Scoping}
The store, $\sigma$, consists of a list of lists of mappings from identifiers to values. Notice the addition of an auxiliary $\kw{end\_scope}$ operator which removes the current scope. A macro literal steps to a macro value by first stepping it's input types until they're all values before substituting all of the values in the store within it's body (forming a closure). Calling a macro value pushes a new empty mappings list onto the store, places the macro's expressions and a $\kw{end\_scope}$ operator onto the stack, enabling local identifiers.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}    
        % Closures capture lexically scoped variables via substitution 
        \Rule{LET-1}
            {v \val
                \\ \sigma = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n] ]
                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n, x_{n+1} \step v_{n+1}] ]}
            {\sigma; v ~ \kweid{x} = \step \sigma'; \varepsilon}

        \Rule{MACRO-1}
            {\tau^* \step \tau^{\prime*}}
            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^{\prime*}}{e^*}}

        \Rule{MACRO-2}
            {\uplus^{|\sigma|}_{i = 1} \sigma(i) = [x_1 \step v_1 ~ ... ~ x_n \step v_n]
                \\ \tau_i \val~ \forall~ \tau_i \in \tau*}
            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n] e^*} \val}

        \Rule{CALL-1}
            {\sigma' = \sigma \cup [~] }
            {\sigma; \kwmacro{\tau^*}{e^*} \textbf{@} \step \sigma'; e^* \kw{end\_scope}}

        \Rule{CALL-2}
            {\sigma' = \text{$\sigma$ but with the last list of mappings removed}}
            {\sigma; \kw{end\_scope} \step \sigma'; \varepsilon} 
    \end{mathpar}  
}

\subsection{Others}
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}

        % Tuple literals evaluation
        \Rule{TUPLES-1}
            {e^* \step e^{*\prime}}
            {( e^* ) \step ( e^{*\prime} )}
        \Rule{TUPLES-2}
            {v_i \val ~ \forall v_i \in v*}
            {(v*) \val}
        \and

        % unpack operator
        \Rule{TUPLES-3}
            {v_i \val ~ \forall~ i \in [1, n]}
            {\kwunpack{(v_1 ... v_n)} \step v_1 ... v_n}
        \Rule{TUPLES-4}
            {\strut}
            {\kwunpack{()} \step \varepsilon}
        \and
        \Rule{TUPLES-5}
            {\tau_i \val ~ \forall~ i \in [1, n]}
            {\kwunpack{(\tau_1 ... \tau_n)} \step \tau_1 ... \tau_n}
        \Rule{TUPLES-6}
            {\strut}
            {\kwunpack{\kwUnit} \step \varepsilon}
        \and

        % math
        \Rule{ADD}
            {v_1 \val \\ v_2 \val \\ v_3 = v_1 + v_2}
            {v_1~ v_2~ + \step v_3}
        \Rule{MUL}
            {v_1 \val \\ v_2 \val \\ v_3 = v_1 * v_2}
            {v_1~ v_2~ \times \step v_3}
        \and

        % ternary
        \Rule{TERN-1}
            {v_1 \val \\ v_2 \val}
            {\kwtern{\kwnumlit{0}{\kwInt}}{v_1}{v_2} \step v_2}
        \Rule{TERN-2}
            {v_1 \val \\ v_2 \val \\ n \neq 0}
            {\kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2} \step v_1}
        \and

        % typeof operator
        \Rule{TYPE-1}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \step \tau} \and
        % Move this to type rules
        % \Rule{Type Operator}
        %     {\tau^* = \tau_1 ~...~ \tau_n \\
        %      v^* = \etyped{v_1}{\tau_1} ~...~ \etyped{v_1}{\tau_1} \\
        %      m = \kwmacro{\tau^*}{e^*} \\
        %      m \val }
        %     {\kwType{m} \step \kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}
    \end{mathpar}
}

\section{Typing}

\subsection{Math and Logic}
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{T-1}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwInt}}{\kwInt}}
        \and \Rule{T-2}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwLong}}{\kwLong}}
        \and \Rule{T-3}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwFloat}}{\kwFloat}}
        \and \Rule{T-4}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwDouble}}{\kwDouble}}
        \and \Rule{T-5}
            {\typed{\ctx}{e_1}{\tau} 
                \\ \typed{\ctx}{e_2}{\tau} 
                \\ \tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\typed{\ctx}{e_1~ e_2~ +}{\tau}}
        \and \Rule{T-6}
            {\typed{\ctx}{e_1}{\tau} 
                \\ \typed{\ctx}{e_2}{\tau} 
                \\ \tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\typed{\ctx}{e_1~ e_2~ \times}{\tau}}
        \and \Rule{T-7}
            {\typed{\ctx}{e_1}{\kwInt}
                \\ \typed{\ctx}{e_2}{\tau}
                \\ \typed{\ctx}{e_3}{\tau}}
            {\typed{\ctx}{\kwtern{e_1}{e_2}{e_3}}{\tau}}
    \end{mathpar}
}
\end{document}
